---
title: "Supplementary Information for: Mathematical characterization of fractal river networks"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: false
    number_sections: false
fontsize: 11pt
mainfont: "Times New Roman"
knit: (function(inputFile, encoding) {
        rmarkdown::render(inputFile,
                          encoding = encoding, 
                          output_dir = "document_output")
      })
bibliography: references.bib
csl: https://www.zotero.org/styles/nature
header-includes:
  - \usepackage{caption}
  - \captionsetup{width=1.3\linewidth, labelfont={bf}}
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{=tex}
\def\figurename{Figure S\hspace{-.4ex}}
\def\tablename{Table S\hspace{-.4ex}}
```
\newpage

| Where      | Original statement                                                                                                                                                                                                                                               | Flaw type                                                                                                                                                                                                                                                                                             |
|------------------|---------------------------|---------------------------|
| p.1        | Here we show that an alleged property of such random networks (branching probability) is a scale-dependent quantity that does not reflect any recognized metric of rivers' fractal character, and hence cannot be a driver of ecological dynamics.               | **Incorrect/Invalid/Improper.** This conclusion rests on the author's misuse of the term "scale invariance" and the flawed analysis with inconsistent units across real rivers.                                                                                                                       |
| p.1        | Moreover, we show that random networks lead to biased estimates of population stability and persistence, while only Optimal Channel Networks yield estimates comparable to real rivers.                                                                          | **Inconclusive.** The author's analysis of metapopulation simulations is fundamentally flawed.                                                                                                                                                                                                        |
| p.2 -- p.3 | *...the former definition of branching probability only holds in the context of the generation of a synthetic random network; it is in fact improper to refer to a "probability"...*                                                                             | **Incorrect.** In real rivers, branching probability has been defined as a cumulative probability distribution of an exponential distribution [@teruiMetapopulationStabilityBranching2018; @teruiEmergentDualScaling2021]                                                                             |
| p.3        | *...*$p_r$ *does not identify the inherent...branching character of a given river network in relation to other river networks.*                                                                                                                                  | **Invalid.** The authors did not non-dimensionalize $p_r$ to estimate the self-similar structure of scale invariant object, which is unique to each river.                                                                                                                                            |
| p.3        | *...by extracting different river networks at various scales...one observes that rivers that look more "branching" (i.e., have higher* $p_r$*) than others for a given* $A_T$ *value can become less "branching" for a different* $A_T$ *(Fig. 3).*              | **Invalid/Improper.** The authors compared incomparable values as the unit of $p_r$ varies by watershed.                                                                                                                                                                                              |
| p.3        | *We therefore conclude that branching probability is a non-descriptive property of a river network, which by no means describes its inherent branching character, and depends on the observation scale*                                                          | **Incorrect/Invalid/Improper.** This statement rests on the improper use of the term "scale-invariance". Branching probability will converge to a single value as $A_T \rightarrow 0$ once properly rescaled.                                                                                         |
| p.3        | *Scaling is also crucial...*                                                                                                                                                                                                                                     | **Likely improper.** The term "scaling" seems to be used to describe a property of scale-dependent object. Scaling is the sufficient condition for scale invariance (scaling $\Leftrightarrow$ scale invariance).                                                                                     |
| p.3        | *...which is inconsistent with the alleged role of* $p_r$ *as a scale-invariant property of river networks.*                                                                                                                                                     | **Invalid/Incorrect/Improper.** This statement build upon the author's misuse of the term "scale invariance"                                                                                                                                                                                          |
| p.3        | *...using the ecological definition of* $l$*...to discretize a real river network into* $N$ *nodes...is problematic.*                                                                                                                                            | **Contradicting.** The authors discretized a real river network into $N$ nodes using the ecological definition of $l$ in their metapopulation simulations.                                                                                                                                            |
| p.4        | Figure 3a and b                                                                                                                                                                                                                                                  | **Invalid/Improper.** The unit of y- and x-axis is not consistent across watersheds.                                                                                                                                                                                                                  |
| p.6        | *Paragraph starting with "However, we need to warn against the conclusion that river networks with higher values of* $p_r$*...are inherently associated with higher metapopulation stability..."*                                                                | **Invalid/Improper.** The statements in this paragraph rests on the results from the flawed analysis (inconsistent units among rivers) made by the authors.                                                                                                                                           |
| p.6        | Figure 6                                                                                                                                                                                                                                                         | **Invalid/Improper.** Metapopulation metrics of real rivers are not comparable because the authors assumed different population scale $l$ and average dispersal distance $\alpha$ by watershed (i.e., different species by watershed)                                                                 |
| p.6 -- p.7 | ...if different river networks spanning different catchment areas (say, in km$^2$) are compared, all of them extracted from the same DEM (same $l$ and same $A_T$ in km$^2$), then the larger river network will appear more branching (i.e., have larger $p_r$) | **Invalid.** Such a correlation was not found in the re-analysis of the author's data. Note that the lack of correlation between branching probability and watershed area $A$ has already been found in previous studies [@teruiMetapopulationStabilityBranching2018; @teruiEmergentDualScaling2021]. |

: Incorrect, invalid, improper, or contradicting statement in the maintext of Carraro and Altermatt 2022 [@carraroOptimalChannelNetworks2022]. See the maintext for detailed discussion.

\newpage

```{r fig-pr, fig.cap="Log-log plot substantiates the power-law scaling of dimensional branching ratio $p_r$ along the axis of observation scale $A_T$. Panels distinguish rivers, with those colored indicating rivers highlighted in the original article [@carraroOptimalChannelNetworks2022]. Individual data points are shown in dots, whose transparency is proportional to the number of links $N_L$ (i.e., sample size). Lines are predicted values from the robust regression model M0 (i.e., the expected value of dimensional branching ratio)."}


knitr::include_graphics(here::here("output/figure_pr_facet.pdf"))
```

\newpage

```{r fig-prp, fig.cap="The properly rescaled branching ratio $\\bar{p_r}$ converges to values unique to each river as $A_T \\rightarrow 0$. Panels distinguish rivers, with those colored indicating rivers highlighted in the original article [@carraroOptimalChannelNetworks2022]. Individual data points are shown in dots, whose transparency is proportional to the number of links $N_L$ (i.e., sample size). Lines are predicted values from the robust regression model M0 (i.e., the expected value of dimensional or rescaled branching ratio)."}


knitr::include_graphics(here::here("output/figure_prp_facet.pdf"))
```

\newpage

# References
